Visualizing Contingency Tables

### 2006-09-05 trunk/vcd/paper/Handbook/hsc.Rnw
index final touches
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@556 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\usepackage{Sweave,hyperref,color,graphicx,makeidx,multicol} 
\usepackage[bottom]{footmisc}% places footnotes at page bottom

@@

\index{Categorical data} Categorical data analysis is typically based on two- or
higher-dimensional \index{contingency tables} contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. Especially
mosaic and sieve plots have been extended over the last two decades, 
and implementations exist in many statistical environments.

@@

Throughout this section, our examples will be based on
the \index{Data!hospital} 
\data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.

@@

\noindent 
Due to the three-dimensional nature of human color
perception \cite{vcd:Mollon:1995}, 
it has been common to specify colors using the three primaries
red, green, and blue (RGB colors), especially for computer devices. 
\index{RGB colors}
The appearance of the on-screen colors is affected by the device
characteristics: For example, the intensity $I$ of a primary 
on a particular device follows the rule $I=L^\gamma$, with $L$
the value for the primary color, and $\gamma$ device-dependent 
(but typically close to 2.2). Therefore, a first caveat is that if colors are to
appear identical on different devices, their gamma characteristics
have to be taken into account.

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on \index{HSV colors}
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette \index{qualitative color palette}
(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
\index{diverging color palette}
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminance better 
contrasts can be achieved.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}),
using a \index{permutation test} permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data \index{Data!arthritis}
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improvement}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

As an example, consider the well-known UCB admissions data
\index{Data!UCB Admissions}
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admission---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

@@

As a four-way example, we use the punishment data \cite{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children (see Table~\ref{tab:punish}).      
\index{Data!Danish corporate punishment}
The data consist of four variables: Attitude is a binary variable
indicating whether a person approves moderate punishment of children
(``moderate''), or refuses any punishment of children (``no'').
Memory indicates whether the person recalls having been
punished as a child. Education indicates the
highest level of education (elementary, secondary, high). Finally, 
age indicates the age group in years (15--24, 25--39, $40+$).

>>>>

\usepackage{Sweave,color,hyperref,graphicx,makeidx,multicol} 
\usepackage[bottom]{footmisc}% places footnotes at page bottom

@@

\index{catgorical data} Categorical data analysis is typically based on two- or
higher-dimensional \index{contingency tables} contingency tables, cross-tabulating the
co-occurences of levels of nominal and/or ordinal data. 
In order to explain these, statisticians
typically look for (conditional) independence structures using common
methods such as independence tests and
log-linear models. One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics). Whether the
task is purely exploratory or model based, techniques such as mosaic, sieve, and
association plots offer good support for visualization. Especially
mosaic and sieve plots have been extended over the last two decades, 
and implementations exist in many statistical environments.

@@

Throughout this section, our examples will be based on
the \index{data!hospital} 
\data{hospital} data \cite{vcd:wing:1962} given in Table~\ref{tab:hospital}.

@@

\noindent 
Due to the three-dimensional nature of human color
perception \cite{vcd:Mollon:1995}, 
it has been common to specify colors using the three primaries
red, green, and blue (RGB colors), especially for computer devices. 
\index{color space!RGB}
The appearance of the on-screen colors is affected by the device
characteristics: For example, the intensity $I$ of a primary 
on a particular device follows the rule $I=L^\gamma$, with $L$
the value for the primary color, and $\gamma$ device-dependent 
(but typically close to 2.2). Therefore, a first caveat is that if colors are to
appear identical on different devices, their gamma characteristics
have to be taken into account.

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on \index{color space!HSV}
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Figure~\ref{fig:colors} (left) 
showing a qualitative color palette \index{color palette!qualitative}
(colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
\index{color palette!diverging}
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Figure~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
\index{color space!CIELAB} \index{color space!CIELUV}
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
\index{color space!HCL}
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Figure~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Figure~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminance better 
contrasts can be achieved.

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference \index{conditional inference}
(or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from (\ref{eq:H0}),
using a \index{permutation test} permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data \index{data!arthritis}
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Table~\ref{tab:arthritis} for the female patients):
\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rrr|r|}
\hline
 & \multicolumn{4}{c|}{\textbf{Improvement}}\\
\textbf{Treatment} & None & Some & Marked & $\Sigma$ \\
\hline
Placebo & 19 & 7 & 6 & 32 \\
Treatment & 6 & 5 & 16 & 27 \\\hline
$\Sigma$ & 25 & 12 & 22 & 59 \\
\hline
\end{tabular}
\caption{The \data{arthritis} data (female patients).}
\label{tab:arthritis}
\end{center}
\end{table}
Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

As an example, consider the well-known UCB admissions data
\index{data!UCB admissions}
\cite{vcd:Bickel+Hammel+O'Connell:1975} on applicants, classified by
admission and gender, to graduate school at the University of California
in Berkeley for the six largest departments in 1973. The `flattened'
contingency table, putting department into the columns and
admission---nested in gender---into the rows, is given by
Table~\ref{tab:ucb}. The table aggregated over all departments gives
wrong evidence of gender bias, i.e., higher admission rates for male
students.

@@

As a four-way example, we use the punishment data \cite{vcd:andersen:1991} 
from a study of the Gallup Institute in Denmark in 1979 about
the attitude of a random sample of 783 persons towards corporal
punishment of children (see Table~\ref{tab:punish}).      
\index{data!Danish corporal punishment}
The data consist of four variables: Attitude is a binary variable
indicating whether a person approves moderate punishment of children
(``moderate''), or refuses any punishment of children (``no'').
Memory indicates whether the person recalls having been
punished as a child. Education indicates the
highest level of education (elementary, secondary, high). Finally, 
age indicates the age group in years (15--24, 25--39, $40+$).
